---
title: "NCAA Men's March Madness Project"
subtitle: "Proposal"
author: 
  - name: "Matt Osterhoudt"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project will examine past NCAA Men's March Madness data in order to develop key insights and predictions utilizing data visualization through plot creation/interpretation"
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
---

```{r}
#| label: load-pkgs
#| message: false


# install and load packages
if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               janitor)

devtools::install_github("tidyverse/dsbox")

library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

team_results <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-03-26/team-results.csv') |>
  clean_names()
public_picks <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/main/data/2024/2024-03-26/public-picks.csv') |>
  clean_names()


view(team_results)
view(public_picks)

```

A brief description of your dataset including its provenance, dimensions, etc. as well as the reason why you chose this dataset.

Make sure to load the data and use inline code for some of this information.

## Questions

1.  Does a team's overall popularity have any correlation to its success in March Madness?
2.   How do public predictions vary across the tournament based on experienced teams?

## Analysis plan

-   Question 1
    -   <div>

        1.  Using the team_results data, I will be using the variables wins (w), win percentage (win), and other indicators of tournament advancement (r64, champ, etc.).
        2.  Using the public_picks data, I will use the cross reference the data from team_results_data and compare the percent of people who picked the team to win finals, final 4, etc. All of those (finals, f4, e8, etc.) are variable names as a percent character.
        3.  I will clean the data by converting character columns to numeric, and also merge both data sets in order to match their round-specific data points for the visualization.
        4.  I will create a new variable that determines the "overall" public popularity. I can do this by creating a new variable that computes the mean of pick percentages. Alternatively, I can compute a weighted score using a mathematical point model system that weighs later rounds as more valuable.
        5.  Finally, I will develop the visualization. I have not decided yet, but I am leaning towards using a histogram, box plot, or scatter plot with a linear regression model.

        </div>

<!-- -->

-   Question 2
    -   <div>

        1.  Using the team_results data, I will convert the amount of games played by teams into a categorical variable. This variable will be called "Experience level" and will be categorized as low, high, and medium (I may adjust this later).
        2.  Using the public_picks data, I will revisit the round percentage popularity and compute the variability (using range, standard deviations, mean, etc.)
        3.  Using the new variables and data, merge the data sets and prepare for visualization.
        4.  Utilize statistical testing and plot data to draw conclusions and interpretation

        </div>
